---
title: "Basic ggplots in R"
author: "Clarke van Steenderen"
date: "2025-02"
output:
  html_document:
    toc: true
    toc_depth: 3
    toc_float: true
---

```{r setup, include=FALSE, message=FALSE, warning=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Using ggplot to create basic plots

Here, we will use the popular **iris** dataset, collected by Edgar Anderson in 1936, to illustrate how ggplot can be used to create elegant and attractive graphics. It contains sepal and petal measurements for 50 individuals from three *Iris* species; namely *I. setosa*, *I. versicolor*, and *I. virginica*. 

Check the R [dataset](https://stat.ethz.ch/R-manual/R-devel/library/datasets/html/00Index.html) package for other interesting data sets to tinker with.

**Some notes on syntax:**

* **package_name::** specifies the package name, where the two colons access all the functions associated with that package
* Hashes (#) denote comments, that are not read by R. Use these to annotate your code
* In ggplot, the plus sign (+) means that you are adding additional "layers" to your plot
* An equal sign (=) is synonymous with the arrow (<-). You can use either to assign names to objects
* Dollar signs ($) are like keys that access features of an object, such as columns or lists
* Use a question mark (?) to access the help file for a particular package (e.g. **?ggplot**)

**Let's get started!**

```{r echo=TRUE, warning=FALSE, message=FALSE}
if (!require("pacman"))
  install.packages("pacman")
pacman::p_load(datasets, xlsx, janitor, ggplot2, Rmisc, dplyr, ggpubr)

# access the embedded dataset
iris_data = datasets::iris
iris_data = janitor::clean_names(iris_data)

# for practice, write this data to the project folder as a .csv file
write.csv(iris_data, file = "data/iris_data.csv", quote = FALSE, row.names = FALSE)

# now read it back in 
iris_data_readin = read.csv("data/iris_data.csv")

# now as an Excel file
xlsx::write.xlsx(iris_data, file = "data/iris_data.xlsx", 
                 sheetName = "iris_data", row.names = FALSE)

iris_data_readin_xlsx = xlsx::read.xlsx(file = "data/iris_data.xlsx", sheetIndex = 1)

# make iris species a factor/grouping variable
iris_data$species = as.factor(iris_data$species)
levels(iris_data$species)

# summarySE function gets summary statistics for a chosen variable across groups
petal_summary = Rmisc::summarySE(data = iris_data, measurevar = "petal_length", 
                                 groupvars = "species")

```

There are a number of ways in R to refer to a particular column in a data set. Have a look a the code below, where we can access the **sepal_width** column in five different ways:

```{r eval=FALSE, warning=FALSE, message=FALSE}

# using the dollar operator - I find this the most intuitive
iris_data$sepal_width

# this says "access the values in all the rows in column 2"
iris_data[, 2]

# a few other approaches to do exactly the same thing
iris_data[[2]]
iris_data[["sepal_width"]]
iris_data[, "sepal_width"]

# using the dplyr package:

# select
dplyr::select(iris_data, sepal_width)

# pull
iris_data %>%
  dplyr::pull(sepal_width)

```

Perhaps you would like to access the petal lengths for rows 1 to 5 only. Here's how you can do this: 

```{r echo=TRUE, warning=FALSE, message=FALSE}

iris_data[1:5, 3]

# or using dplyr
iris_data %>%
  dplyr::slice(1:5) %>%   # Select rows 1 to 5
  dplyr::pull(3) # pull out the third column only

# Have a look at a summary of a particular subset of your data.
# Here we'll look at sepal lengths:
summary(iris_data$sepal_length)

# Take a quick glance at the structure of the dataset:
str(iris_data)

# Check the data type:
class(iris_data)

# Have a precursory glance at the first few rows:
head(iris_data)

# use the glimpse function in the dplyr package:
dplyr::glimpse(iris_data)

```

Note that there are always two spaces for values in the square brackets. The first value refers to the row number, and the second is the column number. A comma always separates these values. If there is no value before or after the comma, R assumes that you want all the available values. E.g. iris_data[,] means "get the values across all rows and columns". Read the documentation for the ``dplyr`` package, as there are so many cool functions to make your life easier!

## ggplot
### Using geom_point()

geom_point() generates dot plots, like this:

```{r echo=TRUE, warning=FALSE, message=FALSE}

ggplot2::ggplot(data = iris_data, aes(x=sepal_length, y=petal_length, 
                                      fill=species)) +
  geom_point(shape = 21) +
  scale_fill_manual(values = c("yellow", "lightblue", "red")) +
  xlab("Sepal length (cm)") +
  ylab("Petal length (cm)") +
  theme_classic() 

# Let's add trendlines
ggplot2::ggplot(data = iris_data, aes(x=sepal_length, y=petal_length, 
                                      fill=species)) +
geom_point(shape = 21) +
  scale_fill_manual(values = c("yellow", "lightblue", "red")) +
  xlab("Sepal length (cm)") +
  ylab("Petal length (cm)") +
  theme_classic() +
  stat_smooth(method = "lm", formula = y~x, geom = "smooth", 
              alpha = 0.2, color = "black", linewidth = 0.4, se = TRUE)

```

Perhaps we want to create box plots for the three species, showing an overall summary. Below, we can use **stat_summary()** to add points for the mean values for each species. We can either set the colours ourselves for each species using **scale_fill_manual()**, or we can use default colour palettes in R, using **scale_fill_brewer()**:

```{r echo=TRUE, warning=FALSE, message=FALSE}
ggplot2::ggplot(data = iris_data, aes(x = species, y=petal_length, 
                                      fill=species)) +
  geom_boxplot(alpha = 0.5) +
  scale_fill_brewer(palette="Set1") +
 # scale_fill_manual(values = c("yellow", "lightblue", "red")) +
  stat_summary(fun = mean, geom = "point", shape = 18, size = 5, fill="black") +
  theme(legend.position="none") +
  xlab("Species") +
  ylab("Petal length (cm)") +
  theme_classic() +
  coord_flip()
```

Explore the different **geom** options in the ggplot package - there are quite a few! Perhaps we want to change our boxplots above into violin or dot plots. This is easy, as we can use the same code as above, but just change **geom_box()** into **geom_violin()**:

```{r echo=TRUE, warning=FALSE, message=FALSE}
ggplot2::ggplot(data = iris_data, aes(x = species, y=petal_length, 
                                      fill=species)) +
  geom_violin(alpha = 0.5) +
  scale_fill_brewer(palette="Set1") +
 # scale_fill_manual(values = c("yellow", "lightblue", "red")) +
  stat_summary(fun = mean, geom = "point", shape = 18, size = 5, fill="black") +
  theme(legend.position="none") +
  xlab("Species") +
  ylab("Petal length (cm)") +
  theme_classic() +
  coord_flip()
```

Or dots:

```{r echo=TRUE, warning=FALSE, message=FALSE}
ggplot2::ggplot(data = iris_data, aes(x = species, y=petal_length, 
                                      fill=species)) +
  geom_dotplot(binaxis = "y", stackdir = "center") +
  scale_fill_brewer(palette="Set1") +
 # scale_fill_manual(values = c("yellow", "lightblue", "red")) +
  theme(legend.position="none") +
  xlab("Species") +
  ylab("Petal length (cm)") +
  theme_classic() 
```


Here, we create a subset for each *Iris* species, using the **filter()** function in the **dplyr** package:

```{r echo=TRUE, warning=FALSE, message=FALSE}
setosa = dplyr::filter(iris_data, species=="setosa")
versicolor = dplyr::filter(iris_data, species=="versicolor")
virginica = dplyr::filter(iris_data, species=="virginica")
```

Now we can run a quick correlation test to see if there is a significant correlation between sepal and petal length in *I. versicolor*:

```{r echo=TRUE, warning=FALSE, message=FALSE}
cor.test(versicolor$sepal_length, versicolor$petal_length, method = "pearson")
```

Here we see a significant correlation (p < 0.05), with a correlation coefficient of r = 0.75. Typically, a value greater than 0.7 is considered to be strong. 

Let's plot only *I. versicolor*, using **dplyr::filter()**:

```{r echo=TRUE, warning=FALSE, message=FALSE}

ggplot2::ggplot(data = versicolor, 
                aes(x=sepal_length, y=petal_length)) +
  geom_point(shape = 21, fill = "lightblue") +
  xlab("Sepal length (cm)") +
  ylab("Petal length (cm)") +
  theme_classic()

```

Let's run a quick linear model, where we check whether there is a relationship between petal and sepal length:

```{r echo=TRUE, warning=FALSE, message=FALSE}
versicolor_lm = lm(data = versicolor, petal_length~sepal_length)
summary(versicolor_lm)
```
If we plot the trend line for this linear regression, the intercept = 0.19, and the gradient = 0.69. The R^2 value is 0.56 (p < 0.05). Let's add this line to our dot plot, with a confidence interval band:

```{r echo=TRUE, warning=FALSE, message=FALSE}

ggplot2::ggplot(data = versicolor, 
                aes(x=sepal_length, y=petal_length)) +
  geom_point(shape = 21, fill = "lightblue") +
  xlab("Sepal length (cm)") +
  ylab("Petal length (cm)") +
  theme_classic() +
  stat_smooth(method = "lm", formula = y~x, geom = "smooth", 
              alpha = 0.2, color = "black", linewidth = 0.4, se = TRUE)
# set se = FALSE to remove the confidence interval band
```

We can also add this line using the **geom_abline()** function:


```{r echo=TRUE, warning=FALSE, message=FALSE}

coeff = coefficients(versicolor_lm)
intercept = coeff[1]
gradient = coeff[2]

ggplot2::ggplot(data = versicolor, 
                aes(x=sepal_length, y=petal_length)) +
  geom_point(shape = 21, fill = "lightblue") +
  # add the abline
  geom_abline(intercept = intercept, slope = gradient, linewidth = 0.5, 
              color = "red", linetype = "dashed") +
  xlab("Sepal length (cm)") +
  ylab("Petal length (cm)") +
  theme_classic() +
  # the stat_regline_equation() function in the ggpubr package 
  # adds the equation as text to the plot
  ggpubr::stat_regline_equation(label.y = 6, aes(label = after_stat(c(eq.label))) ) +
  # the stat_regline_equation() function in the ggpubr package adds the 
  # R-squared value as text to the plot
  ggpubr::stat_regline_equation(label.y = 5.8, label.x = 6.5, 
                                aes(label = after_stat(c(rr.label))) ) 

```

Add facet_wrap() to separate the plots by species:

```{r echo=TRUE, warning=FALSE, message=FALSE}

ggplot2::ggplot(data = iris_data, aes(x=sepal_length, y=petal_length, fill=species)) +
  geom_point(shape = 21) +
  scale_fill_manual(values = c("yellow", "lightblue", "red")) +
  xlab("Sepal length (cm)") +
  ylab("Petal length (cm)") +
  theme_bw() +
  facet_wrap(~species) +
  guides(fill=guide_legend(title="Species")) +
  scale_fill_discrete(labels=c(expression(italic('Setosa')), 'Versicolor', "Virginica")) +
  ggpubr::stat_regline_equation(label.y = 6.5, size = 2.5, 
                                aes(label = after_stat(c(eq.label))) ) +
  ggpubr::stat_regline_equation(label.y = 6.25,
                                aes(label = after_stat(c(rr.label))) ) 

```

### Using geom_boxplot()

The **geom_boxplot()** function creates box-and-whisker summary plots for your data:

```{r echo=TRUE, warning=FALSE, message=FALSE}
ggplot2::ggplot(data = iris_data, aes(x=species, y=petal_length, color=species)) +
  geom_boxplot() +
  scale_color_manual(values = c("forestgreen", "blue", "red")) +
  xlab("Iris species") +
  ylab("Petal length (cm)") +
  theme_bw() +
  theme(legend.position="none") +
  # add points to denote means
  stat_summary(fun = "mean", colour = "black", shape = 4)
```

### Using geom_violin()

```{r echo=TRUE, warning=FALSE, message=FALSE}
ggplot2::ggplot(data = iris_data, aes(x=species, y=petal_length, fill=species)) +
  geom_violin(alpha=0.5) +
  geom_dotplot(binaxis= "y",
               stackdir = "center",
               dotsize = 0.5) +
  scale_fill_manual(values = c("forestgreen", "blue", "red")) +
  xlab("Iris species") +
  ylab("Petal length (cm)") +
  theme_bw() +
  theme(legend.position="none")
```

### Using geom_histogram()

Use **geom_histogram** to generate histogram plots. Set an appropriate bin value:

```{r echo=TRUE, warning=FALSE, message=FALSE}
ggplot2::ggplot(data = iris_data, aes(x=petal_length, fill=species)) +
  geom_histogram(bins = 35) +
  scale_fill_manual(values = c("forestgreen", "blue", "red")) +
  xlab("Petal length (cm)") +
  ylab("Count") +
  theme_bw()
```

Use facet_wrap() here:

```{r echo=TRUE, warning=FALSE, message=FALSE}
ggplot2::ggplot(data = iris_data, aes(x=petal_length, fill=species)) +
  geom_histogram(bins = 35) +
  scale_fill_manual(values = c("forestgreen", "blue", "red")) +
  xlab("Petal length (cm)") +
  ylab("Count") +
  theme_bw() +
  facet_wrap(~species)
```

### Saving your plots as .PNG, .JPG, .PDF, or .SVG

To save a plot, first assign it a name. We will call the last plot we made **petal_length_hist**:

```{r echo=TRUE, warning=FALSE, message=FALSE}
petal_length_hist = ggplot2::ggplot(data = iris_data, 
                                    aes(x=petal_length, fill=species)) +
  geom_histogram(bins = 35) +
  scale_fill_manual(values = c("forestgreen", "blue", "red")) +
  xlab("Petal length (cm)") +
  ylab("Count") +
  theme_bw() +
  facet_wrap(~species)
```

Now we can use the ggsave() function to save it in any format we like. Play around with the dimensions to get an ideal size:

```{r echo=TRUE, warning=FALSE, message=FALSE}
# save as PNG
ggplot2::ggsave("outputs/petal_length.png", plot = petal_length_hist, 
                dpi = 350, width = 8, height = 4)
# save as PDF
ggplot2::ggsave("outputs/petal_length.pdf", plot = petal_length_hist, 
                dpi = 350, width = 8, height = 4)
# save as JPG
ggplot2::ggsave("outputs/petal_length.jpg", plot = petal_length_hist, 
                dpi = 350, width = 8, height = 4)
```
