---
title: "glm_answers4"
author: "Clarke van Steenderen"
date: "2025"
output: html_document
---

```{r setup, include=FALSE, message=FALSE, warning=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Practice exercise 4

```{r, message = FALSE}
if (!require("pacman"))
  install.packages("pacman")
pacman::p_load(xlsx, janitor, ggplot2, Rmisc, dplyr, 
               tidyverse, visreg, glmmTMB, gtsummary, effects, ggeffects,
               patchwork, emmeans, GLMsData)

# Set plot theme
theme_set(theme_classic() +
theme(panel.border = element_rect(colour = "black", fill = NA),
axis.text = element_text(colour = "black"),
axis.title.x = element_text(margin = unit(c(2, 0, 0, 0), "mm")),
axis.title.y = element_text(margin = unit(c(0, 4, 0, 0), "mm")),
legend.position = "none"))
```

```{r}
library(GLMsData)
data("deposit")

head(deposit)

poison.data = deposit %>%
  janitor::clean_names()

poison.data$deposit = as.factor(poison.data$deposit)
poison.data$survived = poison.data$number - poison.data$killed

# get proportions of deaths
poison.data$killed.p = poison.data$killed/poison.data$number

ggplot2::ggplot(data = poison.data, aes(x = deposit, 
                                        y = killed.p, 
                                        fill = insecticide)) +
  geom_bar(alpha = 0.5, colour = "black", stat = "identity", 
           position = position_dodge()) +
  scale_fill_manual(values = c("yellow", "darkorange", "brown")) +
  theme(legend.position = "right")

ggplot2::ggplot(data = poison.data, aes(x = insecticide, 
                                        y = killed)) +
  geom_boxplot(alpha = 0.5, fill = "lightblue")

ggplot2::ggplot(data = poison.data, aes(x = deposit, 
                                        y = killed)) +
  geom_boxplot(alpha = 0.5, fill = "darkred")

```

```{r}
poison.glm.gaus = glm(killed.p ~ insecticide + deposit, data = poison.data,
                   family = gaussian)

DHARMa::simulateResiduals(poison.glm.gaus, plot = TRUE)

poison.glm.bin1 = glm(killed.p ~ insecticide + deposit, data = poison.data,
                   weights = number, # NB
                   family = binomial(link = "logit"))

DHARMa::simulateResiduals(poison.glm.bin1, plot = TRUE)
car::Anova(poison.glm.bin1, type = "II")

# do exactly the same thing, without the weights argument
poison.glm.bin2 = glm(cbind(killed, survived) ~ insecticide + deposit, 
                   data = poison.data,
                   family = binomial(link = "logit"))

DHARMa::simulateResiduals(poison.glm.bin2, plot = TRUE)
car::Anova(poison.glm.bin2, type = "II")

summary(poison.glm.bin1)
summary(poison.glm.bin2)

posthoc.poison.1 = emmeans::emmeans(poison.glm.bin2, 
                                  pairwise ~ insecticide, adjust = "tukey")

posthoc.poison.1$contrasts %>%
  summary(infer = TRUE)

posthoc.poison.2 = emmeans::emmeans(poison.glm.bin2, 
                                  pairwise ~ deposit, adjust = "tukey")

posthoc.poison.2$contrasts %>%
  summary(infer = TRUE)


```

The base case here is insecticide A at deposit value 2. The intercept value therefore means that the proportion of insects killed is **exp(-2.8363)** = 0.05864224 = ~ 6%.

Insecticide B is **exp(0.3060)** = 1.4 times more likely to kill an insect than insecticide A

Insecticide C is **exp(2.9598)** = 19.3 times more likely to kill an insect than insecticide A

Deposit 2.64 is **exp(1.1249)** = 3.1 times more likely to kill an insect than Deposit 2

Deposit 3.48 is **exp(2.2848)** = 9.8 times more likely to kill an insect than Deposit 2

Deposit 8 is **exp(3.8384)** = 46.5 times more likely to kill an insect than Deposit 2
