---
title: "GLM_exercise"
author: "Clarke van Steenderen"
date: "2025"
output: html_document
---

```{r setup, include=FALSE, message=FALSE, warning=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Practice exercise 1

The data set **pollinator_abundance_data.csv** contains pollinator abundances on three different flower species, across three seasons (summer, autumn, and winter).

Run the appropriate GLM on this data, and have a look at whether:

1.  abundance differs across the three flower hosts\
2.  abundance differs across seasons\
3.  season affects abundance differently across the three flower species

Additionally, answer the following:

a.  What is the response variable?\
b.  What is the predictor variable?\
c.  Do we need to include an interaction term in this model?\
d.  Which GLM family was appropriate for this data? Provide the relevant model diagnostic plots.\
e.  Provide box-and-whisker plots to graphically present the data

# Answers

a.  The response variable is `pollinator_abundance`. This is the variable we are focusing on.\
b.  The predictor variables are `flower_sp` and `season`. We want to see how these predictors change our response variable.\
c. Yes, as we want to see whether `season` affects `pollinator_abundance` differently on different `flower_sp`. If we do not include an interaction term, our model will assume that `pollinator_abundance` changes with the same magnitude across the different `flower_sp` over different seasons.\
d. Since we are dealing with count data, the best approaches are the **poisson** or **negative binomial** GLM families, depending on the degree of overdispersion.

Let's begin:

```{r, message = FALSE}
if (!require("pacman"))
  install.packages("pacman")
pacman::p_load(xlsx, janitor, ggplot2, Rmisc, dplyr, 
               tidyverse, visreg, glmmTMB, gtsummary, effects, ggeffects,
               patchwork, emmeans)

# Set plot theme
theme_set(theme_classic() +
theme(panel.border = element_rect(colour = "black", fill = NA),
axis.text = element_text(colour = "black"),
axis.title.x = element_text(margin = unit(c(2, 0, 0, 0), "mm")),
axis.title.y = element_text(margin = unit(c(0, 4, 0, 0), "mm")),
legend.position = "none"))
```

Read in the data, and reorder the season levels:

```{r}
pollinator.data = read.csv("data/pollinator_abundance_data.csv") %>%
  janitor::clean_names()

# relevel order of seasons
pollinator.data$season = factor(pollinator.data$season, 
                                levels = c("Summer", "Autumn", "Winter"))

```

We are working with count data here, and so we will need to fit a Poisson or negative binomial GLM. Let's start by running an ANOVA (Gaussian GLM), just to see how poorly the model fits the data:

```{r}
pollinator.glm.gaus = glm(pollinator_abundance ~ flower_sp + 
                            season + flower_sp:season, data = pollinator.data)

DHARMa::simulateResiduals(pollinator.glm.gaus, plot = TRUE)
```

Now we'll try a Poisson model:

```{r}

pollinator.glm.pois = glm(pollinator_abundance ~ flower_sp + 
                          season + flower_sp:season, family = "poisson",
                          data = pollinator.data)

DHARMa::simulateResiduals(pollinator.glm.pois, plot = TRUE)

DHARMa::testZeroInflation(pollinator.glm.pois)
```

The boxplots look better than the Gaussian model, but the QQ plot shows that the points are not falling along the y = x line, that the KS test is significant, that there is significant overdispersion and outliers.

Let's fit a negative binomial model, to account for overdispersion:

```{r}

# fit a model with just one predictor: flower species
pollinator.glm.negbin.1 = glmmTMB::glmmTMB(pollinator_abundance ~ flower_sp,
                          family = nbinom2(link = "log"),
                          data = pollinator.data)

# fit another model with both flower species and season
pollinator.glm.negbin.2 = glmmTMB::glmmTMB(pollinator_abundance ~ flower_sp + season,
                          family = nbinom2(link = "log"),
                          data = pollinator.data)

# check which model is better -> the model with both predictors is best
anova(pollinator.glm.negbin.1, pollinator.glm.negbin.2, test = "Chisq")

# fit a model with both predictors, and an interaction term
pollinator.glm.negbin.3 = glmmTMB::glmmTMB(pollinator_abundance ~ flower_sp + 
                          season + flower_sp:season,
                          family = nbinom2(link = "log"),
                          data = pollinator.data)

# check whether the interaction term improved the model fit -> yes
anova(pollinator.glm.negbin.2, pollinator.glm.negbin.3, test = "Chisq")

# let's get the model diagnostics for model 3
DHARMa::simulateResiduals(pollinator.glm.negbin.3, plot = TRUE)
DHARMa::testZeroInflation(pollinator.glm.negbin.3)
```

This model fits very well. Let's have a look at the significance of our predictor variables, by using a likelihood ratio test (LRT):

```{r}
car::Anova(pollinator.glm.negbin.3, test = "Chisq", type = "III")

# let's get some averages for flower species
Rmisc::summarySE(data = pollinator.data, measurevar = "pollinator_abundance",
                 groupvars = c("flower_sp"))

# let's get some averages for season
Rmisc::summarySE(data = pollinator.data, measurevar = "pollinator_abundance",
                 groupvars = c("season"))

Rmisc::summarySE(data = pollinator.data, measurevar = "pollinator_abundance",
                 groupvars = c("flower_sp"))

# let's get some averages for everything
pollinator.summary = Rmisc::summarySE(data = pollinator.data, 
                 measurevar = "pollinator_abundance",
                 groupvars = c("flower_sp", "season"))

```

Let's make some box plots:

```{r}

# plot flower species
ggplot2::ggplot(data = pollinator.data, 
                aes(x = flower_sp, y = pollinator_abundance)) +
  geom_boxplot(fill = c("red", "green", "blue"), alpha = 0.5) +
  stat_summary(fun = mean, geom = "point", shape = 18, size = 5, fill="black") +
  xlab("Flower species") + 
  ylab("Pollinator abundance")

# plot seasons
ggplot2::ggplot(data = pollinator.data, 
                aes(x = season, y = pollinator_abundance)) +
  geom_boxplot(fill = c("yellow", "darkorange", "brown"), alpha = 0.5) +
  stat_summary(fun = mean, geom = "point", shape = 18, size = 5, fill="black") +
  xlab("Season") + 
  ylab("Pollinator abundance")

# plot all
ggplot2::ggplot(data = pollinator.data, 
                aes(x = flower_sp, y = pollinator_abundance, 
                    fill = season)) +
  geom_boxplot(alpha = 0.5) +
  scale_fill_manual(values = c("yellow", "darkorange", "brown")) +
  theme(legend.position = "right") +
  xlab("Flower species") + 
  ylab("Pollinator abundance")
```

```{r}
# run some post hoc tests to see differences between groups
posthoc.season = emmeans::emmeans(pollinator.glm.negbin.3, 
                                      pairwise ~ season,
                                      adjust = "tukey")

posthoc.flower = emmeans::emmeans(pollinator.glm.negbin.3, 
                                      pairwise ~ flower_sp,
                                      adjust = "tukey")

posthoc.season$contrasts %>%
  summary(infer = TRUE)

posthoc.flower$contrasts %>%
  summary(infer = TRUE)
```

We can also create boxplots showing means, and standard deviations by using the `geom_errorbar()` function in ggplot. Note how you can modify the axis labels and legend title, and how you can group a bar plot by different factors using `facet_wrap()`.

```{r}

# OVERALL PLOT
ggplot2::ggplot(data = pollinator.summary, 
                aes(x = flower_sp, y = pollinator_abundance, fill = season)) +
  geom_bar(stat = "identity", position = position_dodge(), 
           alpha = 0.65, colour = "black") +
  geom_errorbar(aes(ymin = pollinator_abundance - sd, 
                    ymax = pollinator_abundance + sd), 
                position = position_dodge(width = 0.90), width = 0.25) +
  scale_fill_manual(values = c("yellow", "darkorange", "brown")) +
  theme(legend.position = "right") +
  labs(x = "Flower species", y = "Pollinator abundance", fill = "Season") +
  scale_x_discrete(labels = c("flower_sp1" = "Flower Species 1",
                               "flower_sp2" = "Flower Species 2",
                               "flower_sp3" = "Flower Species 3")) 

# GROUP BY SEASON
ggplot2::ggplot(data = pollinator.summary, 
                aes(x = flower_sp, y = pollinator_abundance, fill = season)) +
  geom_bar(stat = "identity", position = position_dodge(), 
           alpha = 0.65, colour = "black") +
  geom_errorbar(aes(ymin = pollinator_abundance - sd, 
                    ymax = pollinator_abundance + sd), 
                position = position_dodge(width = 0.90), width = 0.25) +
  scale_fill_manual(values = c("yellow", "darkorange", "brown")) +
  theme(legend.position = "right") +
  labs(x = "Flower species", y = "Pollinator abundance", fill = "Season") +
  scale_x_discrete(labels = c("flower_sp1" = "Flower Species 1",
                               "flower_sp2" = "Flower Species 2",
                               "flower_sp3" = "Flower Species 3")) +
  # rotate x axis labels 45 degrees
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  facet_wrap(~season)

# GROUP BY FLOWER SPECIES
ggplot2::ggplot(data = pollinator.summary, 
                aes(x = season, y = pollinator_abundance, fill = flower_sp)) +
  geom_bar(stat = "identity", position = position_dodge(), 
           alpha = 0.65, colour = "black") +
  geom_errorbar(aes(ymin = pollinator_abundance - sd, 
                    ymax = pollinator_abundance + sd), 
                position = position_dodge(width = 0.90), width = 0.25) +
  theme(legend.position = "right") +
  labs(x = "Season", y = "Pollinator abundance", fill = "Flower species") +
  scale_fill_manual(values = c("forestgreen", "gold", "darkred"),
                    labels = c("Flower Species 1", "Flower Species 2", 
                               "Flower Species 3")) +
  # rotate x axis labels 45 degrees
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  facet_wrap(~flower_sp,
             labeller = labeller(flower_sp = c("flower_sp1" = "Flower Species 1",
                                               "flower_sp2" = "Flower Species 2",
                                               "flower_sp3" = "Flower Species 3")))

# look at our summary 
pollinator.summary
```

We can thus report that:

There was a significant difference in pollinator abundance recorded on the three flower species (LRT: $\chi^2$ = 440.97, d.f. = 2, p \< 0.001), and a significant difference across seasons (LRT: $\chi^2$ = 179.53, d.f. = 2, p \< 0.001). Additionally, seasonality had a significant effect on the abundance of pollinators across the three different flower species (LRT: $\chi^2$ = 235.54, d.f. = 4, p \< 0.001).

If we have a look at a boxplot, combined with our summary table, we can also add that:

-   Pollinator abundance was highest on flower sp. 3 (28.4 $\pm$ 19.7), and lowest on flower sp. 1 (10.1 $\pm$ 6.8)
-   Pollinator abundance was highest in summer (28.5 $\pm$ 20.9), and lowest in winter (9.1 $\pm$ 5.9), such that winter abundance was 68% lower: (9.1 - 28.5)/(28.5) x 100
-   For flower sp. 1, pollinator abundance was highest in autumn, while abundances were highest in summer for flower sp. 2 and sp. 3
