---
title: "GLM_exercise_2"
author: "Clarke van Steenderen"
date: "2025"
output: html_document
---

```{r setup, include=FALSE, message=FALSE, warning=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Practice exercise 2

```{r, message = FALSE}
if (!require("pacman"))
  install.packages("pacman")
pacman::p_load(xlsx, janitor, ggplot2, Rmisc, dplyr, 
               tidyverse, visreg, glmmTMB, gtsummary, effects, ggeffects,
               patchwork, emmeans, GLMsData)

# Set plot theme
theme_set(theme_classic() +
theme(panel.border = element_rect(colour = "black", fill = NA),
axis.text = element_text(colour = "black"),
axis.title.x = element_text(margin = unit(c(2, 0, 0, 0), "mm")),
axis.title.y = element_text(margin = unit(c(0, 4, 0, 0), "mm")),
legend.position = "none"))
```

Read in the data collected in the 1970s on the smoking habits of 654 youths. Female = 0, male = 1, non-smoker = 0, smoker = 1, FEV = Forced Expiratory Volume (lung capacity). It seems weird that smokers appear to have a higher lung capacity than non-smokers! Split the plots so that they account for age and height:

```{r}
library(GLMsData)
data(lungcap)

str(lungcap)
lungcap$Smoke = as.factor(lungcap$Smoke)

# Plot FEV for smokers and non-smokers
ggplot2::ggplot(data = lungcap, aes(x = Smoke, y = FEV)) +
  geom_boxplot() +
  facet_wrap(~Gender) +
  scale_x_discrete(labels = c("0" = "Non-smoker",
                               "1" = "Smoker")) 

# Plot FEV for smokers and non-smokers
ggplot2::ggplot(data = lungcap, aes(x = Smoke, y = FEV)) +
  geom_dotplot(binaxis = "y", stackdir = "center", dotsize = 0.5, alpha = 0.5) +
  scale_x_discrete(labels = c("0" = "Non-smoker",
                               "1" = "Smoker")) 

# Another presentation
ggplot2::ggplot(data = lungcap, aes(x = Smoke, y = FEV)) +
  geom_jitter(alpha = 0.5) 

# look at age
ggplot2::ggplot(data = lungcap, aes(x = Gender, y = Age)) +
  geom_boxplot() +
  facet_wrap(~Smoke, 
             labeller = labeller(Smoke = c("0" = "Non-smoker", "1" = "Smoker"))) +
  scale_x_discrete(labels = c("0" = "Female", "1" = "Male")) 

# Plot FEV, taking age into account: LM lines
ggplot2::ggplot(data = lungcap, aes(x = Age, y = FEV, colour = factor(Smoke))) +
  geom_point() +
  scale_colour_manual(values = c("darkgrey", "darkred")) +
  geom_smooth(data = subset(lungcap, Smoke == 0), aes(group = 1), 
              method = "lm", linewidth = 1, col = "darkgrey", se = FALSE) +
  geom_smooth(data = subset(lungcap, Smoke == 1), aes(group = 1), 
              method = "lm", linewidth = 1, col = "darkred", se = FALSE) +
  facet_wrap(~Gender)

# Plot FEV, taking age into account: loess lines
ggplot2::ggplot(data = lungcap, aes(x = Age, y = FEV, colour = factor(Smoke))) +
  geom_point() +
  scale_colour_manual(values = c("darkgrey", "darkred")) +
  geom_smooth(data = subset(lungcap, Smoke == 0), aes(group = 1), 
              method = "loess", linewidth = 1, col = "darkgrey", se = FALSE) +
  geom_smooth(data = subset(lungcap, Smoke == 1), aes(group = 1), 
              method = "loess", linewidth = 1, col = "darkred", se = FALSE) +
  facet_wrap(~Gender)

# Plot FEV taking height into account: LM lines
ggplot2::ggplot(data = lungcap, aes(x = Ht, y = FEV, colour = factor(Smoke))) +
  geom_point() +
  scale_colour_manual(values = c("darkgrey", "darkred")) +
  geom_smooth(data = subset(lungcap, Smoke == 0), aes(group = 1), 
              method = "lm", linewidth = 1, col = "darkgrey", se = FALSE) +
  geom_smooth(data = subset(lungcap, Smoke == 1), aes(group = 1), 
              method = "lm", linewidth = 1, col = "darkred", se = FALSE) +
  facet_wrap(~Gender)

# Plot FEV taking height into account: loess lines
ggplot2::ggplot(data = lungcap, aes(x = Ht, y = FEV, colour = factor(Smoke))) +
  geom_point() +
  scale_colour_manual(values = c("darkgrey", "darkred")) +
  geom_smooth(data = subset(lungcap, Smoke == 0), aes(group = 1), 
              method = "loess", linewidth = 1, col = "darkgrey", se = FALSE) +
  geom_smooth(data = subset(lungcap, Smoke == 1), aes(group = 1), 
              method = "loess", linewidth = 1, col = "darkred", se = FALSE) +
  facet_wrap(~Gender)

```

Let's fit a few models:

```{r}
lung.glm.gaus.1 = glm(FEV ~ Smoke, data = lungcap, family = gaussian)
lung.glm.gaus.2 = glm(FEV ~ Smoke + Gender, data = lungcap, family = gaussian)
lung.glm.gaus.3 = glm(FEV ~ Smoke + Gender + Smoke:Gender, 
                      data = lungcap, family = gaussian)

# model 2 is better than 1 -> including gender is NB
anova(lung.glm.gaus.1, lung.glm.gaus.2, test = "Chisq")

# model 3 is not particularly better than model 2 -> the interaction term
# is not that important
anova(lung.glm.gaus.2, lung.glm.gaus.3, test = "Chisq")

lung.glm.gaus.4 = glm(FEV ~ Smoke + Gender + Age, 
                      data = lungcap, family = gaussian)

# model 4 is better than model 2 -> age is also important
anova(lung.glm.gaus.2, lung.glm.gaus.4, test = "Chisq")

# add the interaction term between smoking and age
lung.glm.gaus.5 = glm(FEV ~ Smoke + Gender + Age + Smoke:Age, 
                      data = lungcap, family = gaussian)

# model 5 is better than model 4 -> interaction between smoking and age is NB
anova(lung.glm.gaus.4, lung.glm.gaus.5, test = "Chisq")

# add an interaction term between smoking, gender, and age
lung.glm.gaus.6 = glm(FEV ~ Smoke + Gender + Age + Smoke*Gender*Age, 
                      data = lungcap, family = gaussian)

# model 6 is better than model 5
# adding the interaction between all three variables makes for a better model
anova(lung.glm.gaus.5, lung.glm.gaus.6, test = "Chisq")

# check model diagnostics
DHARMa::simulateResiduals(lung.glm.gaus.6, plot = TRUE)

car::Anova(lung.glm.gaus.6, test = "LR", type = "III")
summary(lung.glm.gaus.6)

Rmisc::summarySE(data = lungcap, measurevar = "FEV", 
                 groupvars = c("Gender", "Smoke"))

```

The summary output suggests that:

* The intercept (FEV) is 0.67 when all predictors are zero (age = 0, non-smoker, and female)

* Smoking (smoker = 1, non-smoker = 0) increases FEV by 2.41 units, compared to a non-smoker (holding all other variables constant)... weird! But, we haven't accounted for age, gender, or height yet, which comes in with the interaction terms

* Gender affects FEV, where FEV is 0.73 units less in males than in females (holding all other variables constant)

* Age affects FEV, where a one year increase in age increases FEV by 0.18 units (holding all other variables constant)

**Interactions**:

* The effect of smoking on FEV differs between males and females, where FEV is 0.59 units less in male compared to female smokers. This isn't significant, however.

* The effect of smoking on FEV differs across ages, where FEV decreases by 0.19 units with every year increase in age [in smokers], compared to non-smokers

* The effect of gender on FEV differs across ages, where FEV increases by 0.11 units more in males than females with every year increase in age

* There is a three-way interaction, suggesting that the effect of smoking on FEV differs between genders, and varies between ages, but this is not significant

Note the sample sizes in this data -> how might this affect our interpretation? There is a lot more data for non-smokers.

We also don't know what the baseline lung capacity was for participants - there might have been some people with a genetic predisposition for a larger capacity, or maybe some people were physically a lot larger than others. These variables have not been accounted for. Aside from smoking or not, we don't know what the unit of measurement was: some people might smoke two packets a day, while others smoke only one or two cigarettes per day. There might also be differences in other underlying health issues that could skew the results; perhaps some people have asthma or other lung-related illnesses, which also is not accounted for.
